Linear Stochastic Approximation- Constant Step-Size and Iterate Averaging

### 2017-05-18 intro.tex
latest

Informally speaking, smaller step-sizes are better for noise rejection and larger step-sizes lead to faster learning, and at the same time, step-sizes that are too large might result in instability of \eqref{eq:lsaintro}. A useful choice in RL has been the diminishing step-sizes \cite{gtd2,gtdmp,borkarbook,konda-tsitsiklis}, where $\alpha_t\ra 0$ such that $\sum_{t\geq 0} \alpha_t=\infty$. An alternate technique, which we call LSA with constant step-size Polyak-Ruppert (CS-PR) averaging, is to run \eqref{eq:lsaintro} as a constant step-size LSA by choosing $\alpha_t=\alpha>0,\forall t\geq 0$, and output the average $\thh_t\eqdef\frac{1}{t+1}\sum_{i=0}^t \theta_i$. In LSA with CS-PR, $\theta_t$ is an internal variable and $\thh_t$ is the output of the algorithm. The idea of averaging in stochastic approximation to improve noise rejection is a technique that goes back to  \citet{ruppert} and \citet{polyak-judisky}. Intuitively, the CS-PR technique can be seen to address the learning vs noise rejection trade-off, where the constant step-size takes care of the learning  and the averaging of the iterates takes care of the noise rejection.

@@

In this paper, we study LSA with CS-PR (thereby extending prior work by \citet{bach} from SGD to general LSA) where data $(b_t,A_t),t\geq 0$ is $i.i.d$ with some underlying distribution $P$ such that the variance is bounded. An additional restriction we place on $P$ is that the matrix $\EEP{A_t}$ is Hurwitz, i.e., all its eigenvalues have positive real parts.

@@

\item \textbf{Instance Dependent Bounds} For a given $P$, we  measure the performance of LSA with CS-PR in terms of the mean square error (MSE) given by $\EEP{\norm{\thh_t-\ts}^2}$. We show there exists an $\alpha_P>0$ such that the mean squared error (MSE) of LSA with CS-PR is at most $O(\frac{1}{t})$, whenever $\alpha\in (0,\alpha_P)$. The MSE can further be broken down to a \emph{bias} term (arising due to the initial condition $\theta_0$) that decays at a rate $O(\frac{1}{t^2})$ and a \emph{variance} term (due to the noise) that decays at at rate $O(\frac{1}{t})$.
\item \textbf{Uniform Step-Sizes:} It is of interest to know, for a given class $\P$ of distributions, whether it is possible to ensure uniform performance $\forall P\in \P$. We also show (via an example) that there is a class $\P$ that does not `admit' a constant step-size $\alpha_{\P}$ that guarantees uniform performance $\forall P\in \P$. This is a negative result in the context of RL. However, we also argue that the subclass (which we denote by $\P_{\text{SPD},B}$) related to SGD for linear regression does `admit' a uniform constant step-size, thereby recovering a part of the result by \citet{bach}.
\item \textbf{Automatic Step-Size:} We argue informally that for any given $P$, a constant step-size can be chosen in an `automated' fashion. This way we argue that the performance LSA with CS-PR is comparable to other popular step-size choices used for LSAs arising in RL.

@@

\paragraph{Implications:}
\paragraph{Organization:}

>>>>

Informally speaking, smaller step-sizes are better for noise rejection and larger step-sizes lead to faster learning, and at the same time, step-sizes that are too large might result in instability of \eqref{eq:lsaintro}. A useful choice in RL has been the diminishing step-sizes \cite{gtd2,gtdmp,konda-tsitsiklis}, where $\alpha_t\ra 0$ such that $\sum_{t\geq 0} \alpha_t=\infty$. An alternate technique, which we call LSA with constant step-size and Polyak-Ruppert (CS-PR) averaging, is to run \eqref{eq:lsaintro} as a constant step-size LSA by choosing $\alpha_t=\alpha>0,\forall t\geq 0$, and output the average $\thh_t\eqdef\frac{1}{t+1}\sum_{i=0}^t \theta_i$. In LSA with CS-PR, $\theta_t$ is an internal variable and $\thh_t$ is the output of the algorithm. The idea of averaging in stochastic approximation to improve noise rejection is a technique that goes back to  \citet{ruppert} and \citet{polyak-judisky}. Intuitively, the CS-PR technique can be seen to address the learning vs noise rejection trade-off, where the constant step-size takes care of the learning  and the averaging of the iterates takes care of the noise rejection.

@@

In this paper, we study LSAs with CS-PR (thereby extending prior work by \citet{bach} from SGD to general LSAs) where the data sequence $(b_t,A_t),t\geq 0$ is $i.i.d$ with some underlying distribution $P$ such that the variance is bounded. An additional restriction we place on $P$ is that the matrix $\EEP{A_t}$ is Hurwitz, i.e., all its eigenvalues have positive real parts.

@@

\item \textbf{Instance Dependent Bounds:} For a given $P$, we  measure the performance of a given LSA with CS-PR in terms of the mean square error (MSE) given by $\EEP{\norm{\thh_t-\ts}^2}$. We show that there exists an $\alpha_P>0$ such that the mean squared error (MSE) %of a given LSA with CS-PR
is at most $O(\frac{1}{t})$, whenever $\alpha\in (0,\alpha_P)$.
%The MSE can further be broken down to a \emph{bias} term (arising due to the initial condition $\theta_0$) that decays at a rate $O(\frac{1}{t^2})$ and a \emph{variance} term (due to the noise) that decays at at rate $O(\frac{1}{t})$.
\item \textbf{Uniform Step-Sizes:} It is of interest to know that for a given LSA with CS-PR and a given class $\P$ of distributions, whether it is possible to ensure uniform performance $\forall P\in \P$. We also show (via an example) that there is a class $\P$ that does not `admit' a constant step-size $\alpha_{\P}$ that guarantees uniform performance $\forall P\in \P$. This is a negative result in the context of RL. However, we also argue that the subclass (which we denote by $\P_{\text{SPD},B}$) related to SGD for linear regression does `admit' a uniform constant step-size, thereby recovering a part of the result by \citet{bach}.
\item \textbf{Automatic Step-Size:} We argue informally that for any given $P$, a constant step-size can be chosen in an `automated' fashion. This way we argue that the performance of LSA with CS-PR is comparable to other popular step-size choices used for LSAs arising in RL.

@@

\paragraph{Implications:} In addition to TD(0), our results directly can be applied to a host of \emph{off-policy} TD algorithms such as GTD/GTD2/TDC with CS-PR. In the case of TDC only asymptotic bounds were known \cite{gtd2} and our results are a first step towards proving finite time bounds. Also, our results show that the GTD class of algorithms guarantee a $O(\frac{1}{t})$ rate for MSE (without use of projections) as opposed to the previous results (using projections) \cite{gtdmp} that provide a $O(\frac{1}{\sqrt{t}})$ rate for this class.
\paragraph{Organization:} The paper is organized as follows

### 2017-05-18 linal.tex
latest

\begin{align*}
\EE{\norm{\eh_t}^2}\geq \EE{\eh^2_t(1)}=\frac{1}{(t+1)^2}(\alpha \lambda_{\min})^{-2}\big[\left(1-(1-\alpha \lambda_{\min})^{t+1}\right)^2 \theta^2_0(1)+ \frac{1}{(t+1)^2}\sum_{s=1}^t\left(1-(1-\alpha \lambda_{\min})^{t+1-s}\right)^2 b^2_s(1) \big]
\end{align*}
Note that in this example, $\rhos{P}=\rhod{P}=2\lambda_{\min} -\alpha \lambda_{\min}^2=\lambda_{\min}(2-\alpha \lambda_{\min})$, and $\norm{\ts}=0$ and $\sigma^2_A=0$. Further, the result follows by noting the fact that noting the fact that $\norm{b_t}^2=b_t(1)^2$ and $\norm{\theta_t}^2=\theta_t(1)^2$.
%\begin{align}
%\EE{\norm{\eh_t}^2}\geq \frac{1}{(t+1)^2}(\alpha \lambda_{\min})^{-2}(1-\alpha \lambda_{\min}^t) \theta^2_0(1)
%\end{align}

>>>>

\begin{align*}
\EE{\norm{\eh_t}^2}\geq \EE{\eh^2_t(1)}\stackrel{(b)}{=}\frac{1}{(t+1)^2}(\alpha \lambda_{\min})^{-2}\big[\left(1-(1-\alpha \lambda_{\min})^{t+1}\right)^2 \theta^2_0(1)+ \frac{1}{(t+1)^2}\sum_{s=1}^t\left(1-(1-\alpha \lambda_{\min})^{t+1-s}\right)^2 b^2_s(1) \big]
\end{align*}
Here $(a)$ and $(b)$ follows from the $i.i.d$ assumption. Note that in this example, $\rhos{P}=\rhod{P}=2\lambda_{\min} -\alpha \lambda_{\min}^2=\lambda_{\min}(2-\alpha \lambda_{\min})$, and $\norm{\ts}=0$ and $\sigma^2_A=0$. Further, the result follows by noting the fact that noting the fact that $\norm{b_t}^2=b_t(1)^2$ and $\norm{\theta_t}^2=\theta_t(1)^2$.
%\begin{align}
%\EE{\norm{\eh_t}^2}\geq \frac{1}{(t+1)^2}(\alpha \lambda_{\min})^{-2}(1-\alpha \lambda_{\min}^t) \theta^2_0(1)
%\end{align}

\paragraph{Proof of \Cref{lm:notwad}}
Fix an arbitrary $\alpha>0$. We show that there exists $P\in \P$ such that $\rho_\alpha(P)<0$.
For $\epsilon \in (0,1/2)$ let $P=(P^V,P^M)$ be the distribution such that $P^M$ is supported on $\{-I,I\}$ and takes on the value of $I$ with probability $1/2+\epsilon$. Then $A_P = 2\epsilon I \succ 0$, hence $P\in \P_1$. Further, $Q_P = I$.
Hence, $\rhos{P} = 4\epsilon-\alpha$. Hence, if $\epsilon<\alpha/4$, $\rho_\alpha(P)<0$.

\paragraph{Proof of \Cref{lm:ppsdbwd}}
Since $\P_{\text{PSD},B}$ is supported on the set of positive semi-definite matrices, we know for any $A\in \R^{\dcd}$ that is PSD, we can consider the SVD of $A$: $A = U \Lambda U^\top$ where $U$ is orthonormal and $\Lambda$ is diagonal with
nonnegative elements. Note that $\Lambda \preceq B\, \I$ and thus $\Lambda^2 \preceq B \Lambda$.
Then for any $x\in \R^d$, $x^\top A^\top A x = x^\top U \Lambda^2 U^\top x \le B x^\top U \Lambda U^\top x = B x^\top A x$.
Taking expectations we find that $x^\top C_P x \le B x^\top A_P x$.
Hence, $\rhos{P_{\text{PSD},B}} = 2 x^\top A_P x - \alpha x^\top C_P x \ge (2- \alpha B ) \,x^\top A_P x $.
Thus, for any $\alpha<2/B$, $\rho_\alpha(P)>0$.

\paragraph{Proof of \Cref{lm:ppsdbna}}
Consider the case when the smallest eigenvalue of $A_P$ is $0$.

### 2017-05-18 uniform.tex
latest

\section{Uniform stepsizes}\label{sec:uniform}
If $\P$ is weakly admissible, then one can choose a step-size $\alpha_{\P}>0$ solely based on the knowledge of $\P$ and
conclude that no matter the joint distribution $P$ of $(b_t,A_t)$, as long as the distribution of $P$ is an element of $\P$, the conclusions of \Cref{th:rate} hold. Further, if $\P$ is admissible then the error bound stated in  \cref{maintheorem} becomes independent of the instance. Hence, an interesting question to investigate is whether a given set $\P$ is (weakly) admissible. We consider some reasonable classes of distributions and test whether or not they are (weakly) admissible.
\begin{definition}\label{def:pclass}
\begin{itemize}[leftmargin=*, before = \leavevmode\vspace{-\baselineskip}]
\item For $B>0$, let $\P_B$ be the set of distributions over $\C_B^{d}\times \C_B^{\dcd}$ where for every $P\in \P_B$ is positive definite.
\item  $\P_{\text{PSD},B}$ be such that  $\P^M_{\text{PSD},B}=\{ P^M \in \P^M_B \,:\,  \supp(P^M)\subset \text{PSD} \}\,$ , where PSD denotes the set of positive semi-definite matrices.
\item  $\P_{\text{PD},B}$ be such that  $\P^M_{\text{SD},B}=\{ P^M \in \P^M_B \,:\,  \supp(P^M)\subset \text{PD} \}\,$ , where PD denotes the set of positive definite matrices.
\end{itemize}
\end{definition}
Note that all distributions are a subset of $\P_B$ and in most applications the knowledge of a suitable $B>0$ is available,and by appropriately normalizing the vectors and matrices involved, we may as well assume that $B=1$ without any loss of generality.
\begin{proposition}\label{notwad}
The set $\P_B$ is not weakly admissible.
\end{proposition}
\begin{proposition}
For any $B>0$, the set $\P_{\text{PSD},B}$ is weakly admissible
and in particular any $0<\alpha < 2/B$ witnesses this.
\end{proposition}
\begin{proposition}
For any $B>0$, the set $\P_{\text{PSD},B}$ is not admissible.
\end{proposition}

@@

\begin{comment}
\begin{proof}
Take any $P\in \P_{\text{PSD}}$ and let $H\sim P$.
Consider the SVD of $H$: $H = U \Lambda U^\top$ where $U$ is orthonormal and $\Lambda$ is diagonal with
nonnegative elements. Note that $\Lambda \preceq B\, \I$ and thus $\Lambda^2 \preceq B \Lambda$.
Then for any $x\in \R^d$, $x^\top H^\top H x = x^\top U \Lambda^2 U^\top x \le B x^\top U \Lambda U^\top x = B x^\top H x$.
Taking expectations we find that $x^\top Q_P x \le B x^\top H_P x$.
Hence, $\rho_\alpha(P) = 2 x^\top H_P x - \alpha x^\top Q_P x \ge (2- \alpha B ) \,x^\top H_P x $.
Thus, for any $\alpha<2/B$, $\rho_\alpha(P)>0$.
\end{proof}
\end{comment}
\todoc[inline]{So this altogether is weaker than what Bach founds. They prove strong admissibility.
I am guessing that for this they use that the loss is special and also that the noise is special.
This would be a good place to somehow incorporate this with a (short?) proof.
}
Thus \Cref{notwad} positive definiteness of $P\in \P_B$ alone is not sufficient for weak admissibility. However, when we impose additional structure in the data in the form of positive semi-definiteness,  we then observe weak admissibility in \Cref{wad}, and at the same time, it is not enough to guarantee admissibility in \Cref{notad}. Finally, when we impose structure in the form of positive definiteness we observe admissibility.
\begin{comment}
\FloatBarrier
\begin{table}[H]
\begin{tabular}{|c|c|c|c|}\hline
Problem Class & Weak Admissible & Admissible &Example \\ \hline
$\P_B$ &\ding{53} &\ding{53} & RL\\\hline
$\P_{\text{PSD},B}$ &\ding{51} &\ding{53} & SGD for Linear Regression\\\hline
$\P_{\text{PD},B}$ &\ding{51} &\ding{51} & SGD for Linear Regression with regularization\\\hline
\end{tabular}
\end{table}
\end{comment}

>>>>

\section{Uniform stepsizes}\label{sec:uniform}
If $\P$ is weakly admissible, then one can choose a step-size $\alpha_{\P}>0$ solely based on the knowledge of $\P$ and
conclude that no matter the joint distribution $P$ of $(b_t,A_t)$, as long as the distribution of $P$ is an element of $\P$, the conclusions of \Cref{th:rate} hold. Further, if $\P$ is admissible then the error bound stated in  \cref{maintheorem} becomes independent of the instance. Hence, an interesting question to investigate is whether a given set $\P$ is (weakly) admissible. We consider some reasonable classes of distributions and test whether or not they are (weakly) admissible.
\begin{definition}\label{def:pclass}
\begin{itemize}[leftmargin=*, before = \leavevmode\vspace{-\baselineskip}]
\item For $B>0$, let $\P_B$ be the set of distributions over $\C_B^{d}\times \C_B^{\dcd}$ where for every $P\in \P_B$ is positive definite.
\item  $\P_{\text{PSD},B}$ be such that  $\P^M_{\text{PSD},B}=\{ P^M \in \P^M_B \,:\,  \supp(P^M)\subset \text{PSD} \}\,$ , where PSD denotes the set of positive semi-definite matrices.
%\item  $\P_{\text{PD},B}$ be such that  $\P^M_{\text{SD},B}=\{ P^M \in \P^M_B \,:\,  \supp(P^M)\subset \text{PD} \}\,$ , where PD denotes the set of positive definite matrices.
\end{itemize}
\end{definition}
Note that all distributions are a subset of $\P_B$ and in most applications the knowledge of a suitable $B>0$ is available,and by appropriately normalizing the vectors and matrices involved, we may as well assume that $B=1$ without any loss of generality.
\begin{proposition}\label{notwad}
The set $\P_B$ is not weakly admissible.
\end{proposition}
\begin{proposition}\label{lm:ppsdbwd}
For any $B>0$, the set $\P_{\text{PSD},B}$ is weakly admissible
and in particular any $0<\alpha < 2/B$ witnesses this.
\end{proposition}
\begin{proposition}\label{lm:ppsdbna}
For any $B>0$, the set $\P_{\text{PSD},B}$ is not admissible.
\end{proposition}

@@

\begin{comment}
\begin{proof}
Take any $P\in \P_{\text{PSD}}$ and let $H\sim P$.
Consider the SVD of $H$: $H = U \Lambda U^\top$ where $U$ is orthonormal and $\Lambda$ is diagonal with
nonnegative elements. Note that $\Lambda \preceq B\, \I$ and thus $\Lambda^2 \preceq B \Lambda$.
Then for any $x\in \R^d$, $x^\top H^\top H x = x^\top U \Lambda^2 U^\top x \le B x^\top U \Lambda U^\top x = B x^\top H x$.
Taking expectations we find that $x^\top Q_P x \le B x^\top H_P x$.
Hence, $\rho_\alpha(P) = 2 x^\top H_P x - \alpha x^\top Q_P x \ge (2- \alpha B ) \,x^\top H_P x $.
Thus, for any $\alpha<2/B$, $\rho_\alpha(P)>0$.
\end{proof}
\end{comment}
\todoc[inline]{So this altogether is weaker than what Bach founds. They prove strong admissibility.
I am guessing that for this they use that the loss is special and also that the noise is special.
This would be a good place to somehow incorporate this with a (short?) proof.
}
Thus \Cref{lm:notwad} positive definiteness of $P\in \P_B$ alone is not sufficient for weak admissibility. However, when we impose additional structure in the data in the form of positive semi-definiteness,  we then observe weak admissibility in \Cref{lm:ppsdwd}, and at the same time, it is not enough to guarantee admissibility in \Cref{lm:ppsdna}.
\begin{comment}
\FloatBarrier
\begin{table}[H]
\begin{tabular}{|c|c|c|c|}\hline
Problem Class & Weak Admissible & Admissible &Example \\ \hline
$\P_B$ &\ding{53} &\ding{53} & RL\\\hline
$\P_{\text{PSD},B}$ &\ding{51} &\ding{53} & SGD for Linear Regression\\\hline
$\P_{\text{PD},B}$ &\ding{51} &\ding{51} & SGD for Linear Regression with regularization\\\hline
\end{tabular}
\end{table}
\end{comment}
